---
title: "mouse"
author: "Naoki Matsumaru"
date: "2024-04-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r reading libraries}
library(tidyverse)
library(Seurat)
library(openxlsx)
library(hdf5r)
library(ggplot2)
library(HGNChelper)
#library(SeuratData)
set.seed(1234)
```
Data acquisition

source:

https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE224398

available on h5 format.

```{r reading h5 files}
raw_WT.data <- Read10X_h5("GSM7021714_C_6m_filtered_feature_bc_matrix.h5")
raw_AD.data <- Read10X_h5("GSM7021717_A_6m_filtered_feature_bc_matrix.h5")
```

To see what is in the h5 file.

```{r}
raw_WT.data %>% class
raw_WT.data %>% dim
raw_WT.data %>% as.data.frame()
```


dgcMatrix is a class of sparse numeric matrices in the compressed, sparse, column-oriented format.


https://github.com/IanevskiAleksandr/sc-type/blob/master/README.md

First, need to annotate cell type

```{r}
raw_WT <- CreateSeuratObject(counts = raw_WT.data, project = "ScRNA-seq of AD mouse")
raw_WT$type <- "WT"
```


```{r}
# normalize data
raw_WT[["percent.mt"]] <- PercentageFeatureSet(raw_WT, pattern = "^MT-")
raw_WT <- NormalizeData(raw_WT)
```


```{r}
raw_WT <- FindVariableFeatures(raw_WT, selection.method = "vst", nfeatures = 3000)
```


```{r}
raw_WT <- ScaleData(raw_WT, features = rownames(raw_WT))
raw_WT <- RunPCA(raw_WT, features = VariableFeatures(object=raw_WT))
```


```{r}
# Check number of PC components (we selected 10 PCs for downstream analysis, based on Elbow plot)
ElbowPlot(raw_WT)
```

```{r}

```


```{r}

raw_AD <- CreateSeuratObject(counts = raw_AD.data, project = "ScRNA-seq of AD mouse")
raw_AD$type<- "AD"
```
```{r}

raw_AD <- NormalizeData(raw_AD)
```
```{r}

raw_AD <- FindVariableFeatures(raw_AD, selection.method = "vst", nfeatures = 3000)
```

# Data scaling

```{r}

raw_AD <- ScaleData(raw_AD, features = rownames(raw_AD))
raw_AD <- RunPCA(raw_AD, features = VariableFeatures(object=raw_AD))


DimHeatmap(raw_WT, dims = 1:3, cells = 500, balanced = T)
```

```{r}
count_list <- list("WT_6m"=raw_WT.data, "AD_6m"=raw_AD.data)
```



## Create Seurat Object
```{r}
mouse_obj <- CreateSeuratObject(count=count_list)
```

To see what is Seurat object

```{r}
mouse_obj@meta.data %>% as.data.frame()
mouse_obj%>% names
Layers(mouse_obj[["RNA"]])
```


```{r}
# load gene set preparation function
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R")
# load cell type annotation function
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")

```



```{r}
# DB file
db_ <- "https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/ScTypeDB_full.xlsx";
tissue = "Brain"
#tissue <- "Immune system" # e.g. Immune system,Pancreas,Liver,Eye,Kidney,Brain,Lung,Adrenal,Heart,Intestine,Muscle,Placenta,Spleen,Stomach,Thymus 

# prepare gene sets
gs_list <- gene_sets_prepare(db_, tissue)

```

```{r}
# check Seurat object version (scRNA-seq matrix extracted differently in Seurat v4/v5)
seurat_package_v5 <- isFALSE('counts' %in% names(attributes(raw_WT[["RNA"]])));
print(sprintf("Seurat object %s is used", ifelse(seurat_package_v5, "v5", "v4")))
```
https://satijalab.org/seurat/archive/v3.0/merge_vignette.html
https://satijalab.org/seurat/archive/v4.3/merge
https://satijalab.org/seurat/articles/essential_commands.html



###QUALITY CHECK###-------------------------------------------------------------

```{r}
raw_WT.data %>% as.data.frame() %>% filter(str_detect(row.names(.), "^MT-"))
raw_AD.data %>% as.data.frame() %>% filter(str_detect(row.names(.), "^MT-"))
```

No MT related genes

```{r}
raw_WT[["percent.mt"]] <- PercentageFeatureSet(raw_WT, pattern = "^MT-")
raw_AD[["percent.mt"]] <- PercentageFeatureSet(raw_AD, pattern = "^MT-")
```


```{r}
VlnPlot(raw_WT, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

No mitochondrial gene.

```{r}
FeatureScatter(raw_WT, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```



```{r}
VlnPlot(raw_AD, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```


```{r}
#FeatureScatter(seurat_AD, feature1 = "nCount_RNA", feature2 = "percent.mt")
FeatureScatter(raw_AD, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```


```{r}
raw_WT <- subset(raw_WT, subset = nFeature_RNA > 200 & nFeature_RNA < 5000 & percent.mt < 30)
raw_AD <- subset(raw_AD, subset = nFeature_RNA > 200 & nFeature_RNA < 3000 & percent.mt < 25)
```


```{r}
###NORMALIZATION###-------------------------------------------------------------
raw_WT <- NormalizeData(raw_WT, normalization.method = "LogNormalize", scale.factor = 10000)
raw_AD <- NormalizeData(raw_AD, normalization.method = "LogNormalize", scale.factor = 10000)
```


```{r}
raw_WT <- FindVariableFeatures(raw_WT, selection.method = "vst", nfeatures = 3000)
raw_AD <- FindVariableFeatures(raw_AD, selection.method = "vst", nfeatures = 3000)
```


```{r, fig.width=13, fig.height=5}
top10 <- head(VariableFeatures(raw_WT), 10)
plot1 <- VariableFeaturePlot(raw_WT)
plot21 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
#plot21
top10 <- head(VariableFeatures(raw_AD), 10)
plot1 <- VariableFeaturePlot(raw_AD)
plot22 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot21 + plot22
```

# Data scaling

```{r}
raw_WT <- ScaleData(raw_WT)
raw_WT <- RunPCA(raw_WT, features = head(VariableFeatures(object=raw_WT), 10000))
DimHeatmap(raw_WT, dims = 1:3, cells = 500, balanced = T)
```

```{r}
raw_AD <- ScaleData(raw_AD)
raw_AD <- RunPCA(raw_AD, features = head(VariableFeatures(object=raw_AD), 10000))
DimHeatmap(raw_AD, dims = 1:3, cells = 500, balanced = T)
```
```{r}
ElbowPlot(raw_WT)
ElbowPlot(raw_AD)
```
```{r}
raw_WT <- FindNeighbors(raw_WT, dims = 1:5)
raw_AD <- FindNeighbors(raw_AD, dims = 1:5)
```
```{r}
raw_WT <- FindClusters(raw_WT, resolution=0.8)
raw_AD <- FindClusters(raw_AD, resolution=0.8)
```
```{r}
raw_WT <- RunUMAP(raw_WT, dims=1:5)
raw_AD <- RunUMAP(raw_AD, dims=1:5)
```

```{r}
DimPlot(raw_WT,reduction="umap")
DimPlot(raw_AD,reduction="umap")

```



```{r}
# extract scaled scRNA-seq matrix
scRNAseqData_scaled <- if (seurat_package_v5) as.matrix(raw_WT[["RNA"]]$scale.data) else as.matrix(raw_WT[["RNA"]]@scale.data)
```


```{r}
es.max = sctype_score(scRNAseqData = scRNAseqData_scaled, scaled = TRUE, gs = gs_list$gs_positive, gs2 = gs_list$gs_negative)
```



```{r}
# merge by cluster
cL_resutls <- do.call("rbind", lapply(unique(raw_WT@meta.data$seurat_clusters), function(cl){
    es.max.cl = sort(rowSums(es.max[ ,rownames(raw_WT@meta.data[raw_WT@meta.data$seurat_clusters==cl, ])]), decreasing = !0)
    head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(raw_WT@meta.data$seurat_clusters==cl)), 10)
}))
sctype_scores <- cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)  

# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] <- "Unknown"
print(sctype_scores[,1:3])
```

```{r}
raw_WT@meta.data$sctype_classification = ""
for(j in unique(sctype_scores$cluster)){
  cl_type = sctype_scores[sctype_scores$cluster==j,]; 
  raw_WT@meta.data$sctype_classification[raw_WT@meta.data$seurat_clusters == j] = as.character(cl_type$type[1])
}

dim_plot_wt <- DimPlot(raw_WT, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'sctype_classification')        
```

```{r}
dim_plot_wt
```


```{r}
# check Seurat object version (scRNA-seq matrix extracted differently in Seurat v4/v5)
seurat_package_v5 <- isFALSE('counts' %in% names(attributes(raw_AD[["RNA"]])));
print(sprintf("Seurat object %s is used", ifelse(seurat_package_v5, "v5", "v4")))

```


```{r}
# extract scaled scRNA-seq matrix
scRNAseqData_scaled <- if (seurat_package_v5) as.matrix(raw_AD[["RNA"]]$scale.data) else as.matrix(raw_AD[["RNA"]]@scale.data)
```


```{r}
es.max = sctype_score(scRNAseqData = scRNAseqData_scaled, scaled = TRUE, gs = gs_list$gs_positive, gs2 = gs_list$gs_negative)
```



```{r}
# merge by cluster
cL_resutls <- do.call("rbind", lapply(unique(raw_AD@meta.data$seurat_clusters), function(cl){
    es.max.cl = sort(rowSums(es.max[ ,rownames(raw_AD@meta.data[raw_AD@meta.data$seurat_clusters==cl, ])]), decreasing = !0)
    head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(raw_AD@meta.data$seurat_clusters==cl)), 10)
}))
sctype_scores <- cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)  

# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] <- "Unknown"
print(sctype_scores[,1:3])
```

```{r}
raw_AD@meta.data$sctype_classification = ""
for(j in unique(sctype_scores$cluster)){
  cl_type = sctype_scores[sctype_scores$cluster==j,]; 
  raw_AD@meta.data$sctype_classification[raw_AD@meta.data$seurat_clusters == j] = as.character(cl_type$type[1])
}

dim_plot_ad <- DimPlot(raw_AD, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'sctype_classification')        
```

```{r}
dim_plot_ad
```
```{r}
target_rownames<- raw_WT@meta.data %>% filter(sctype_classification == "Microglial cells") %>% row.names()

Lgals3_WT <- raw_WT[["RNA"]]$data %>% as.data.frame() %>%  filter(str_detect(row.names(.), "Lgals3")) %>%
              select(all_of(target_rownames))
```


```{r}
target_rownames %>% length
Lgals3_WT %>% t %>% dim
```
```{r}
raw_WT@meta.data  %>% merge(t(Lgals3_WT), by="row.names", all=TRUE) %>% filter(sctype_classification == "Microglial cells")
```



```{r}
target_rownames<- raw_AD@meta.data %>% filter(sctype_classification == "Microglial cells") %>% row.names()

Lgals3_AD <- raw_AD[["RNA"]]$data %>% as.data.frame() %>%  filter(str_detect(row.names(.), "Lgals3")) %>%
              select(all_of(target_rownames))

```

```{r}
WT_Lgals3<- raw_WT@meta.data %>% merge(t(Lgals3_WT), by="row.names", all=TRUE) %>% filter(sctype_classification == "Microglial cells")
AD_Lgals3<- raw_AD@meta.data %>% merge(t(Lgals3_AD), by="row.names", all=TRUE) %>% filter(sctype_classification == "Microglial cells")
```


```{r}
plot_dat <- bind_rows(WT_Lgals3, AD_Lgals3)
```

```{r}
plot_dat
```

```{r}
plot_dat %>% filter(Lgals3>0) %>% ggplot(aes(x=type, y=Lgals3)) + geom_boxplot() + geom_jitter()
```
```{r}
plot_dat %>% filter(Lgals3bp>0) %>% ggplot(aes(x=type, y=Lgals3bp)) + geom_boxplot() + geom_jitter()
```




```{r}
library(cowplot)
plot_grid(dim_plot_wt,dim_plot_ad, nrow=2)
```



```{r}
raw_WT[["RNA"]]

raw_WT$type
```

```{r}
all.equal(colnames(raw_WT.data),colnames(raw_AD.data))
```

```{r}
amga <- CreateSeuratObject(counts=raw_AD.data)
Assays(amga)
```
```{r}
wt_assay <- CreateAssay5Object(counts=raw_WT.data)
amga[["WT"]] <- wt_assay
Assays(amga)
```
```{r}

```


```{r}
df_db<- read.xlsx(db_)
```

```{r}
df_db %>% filter(tissueType=='Brain') %>% select(geneSymbolmore1)
#  filter(geneSymbolmore1 %in% "Lgals3")

df_db %>% filter(tissueType=='Brain') %>%   filter(str_detect(geneSymbolmore1,"Lgals3"))
df_db %>% filter(tissueType=='Brain') %>%   filter(str_detect(geneSymbolmore2,"Lgals3"))

```

## Looking for Lgals3 in the cluster

```{r}
boud_data <- bind_rows(raw_WT.data %>% as.data.frame() %>% filter(str_detect(row.names(.), "Lgals3")), raw_AD.data %>% as.data.frame() %>% filter(str_detect(row.names(.), "Lgals3")))
```
```{r}
boud_data
```
```{r}
raw_WT.data %>% dim()
raw_AD.data %>% dim

```

```{r}
raw_WT.data %>% as.data.frame() %>% filter(str_detect(row.names(.), "Lgals3")) %>% colnames()
raw_AD.data %>% as.data.frame() %>% filter(str_detect(row.names(.), "Lgals3")) %>% colnames()

```



```{r}
raw_WT[["RNA"]]$data %>% as.data.frame() %>% filter(str_detect(row.names(.), "Lgals3"))
```

```{r}
raw_WT[["RNA"]]$data %>% as.data.frame() %>% filter(str_detect(row.names(.), "Lgals3")) %>% select()
```

